---
title: "Getting Started with nestprepper"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with nestprepper}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction
The `nestprepper` package is designed to take messy sea turtle nesting counts and transform them into a standardized Bayesian trend analysis.

## Step 1: Preparing Your Own Data
To use your own data, ensure your CSV file has at least these three columns:
1. **Year**: The nesting season (e.g., 2020)
2. **Site**: The name of the beach (e.g., "Maehama")
3. **Count**: The number of nests recorded.

If your data is currently "Wide" (each beach has its own column), you should pivot it first:

```{r setup, eval=FALSE}
library(nestprepper)
library(tidyr)
library(dplyr)

# Example of pivoting wide data to the format nestprepper needs:
# raw_data <- read.csv("your_data.csv") %>%
#   pivot_longer(cols = -Year, names_to = "Site", values_to = "Count")
```

## Step 2: The QAQC Gatekeeper
Before running the model, use `prep_nesting_data()`. This function will interactively ask you about monitoring effort and flag any extreme outliers (like typos) that might crash a Bayesian model.

```{r qaqc, eval=FALSE}
# This will trigger interactive questions in your console
clean_data <- prep_nesting_data(your_data_object)
```



## Step 3: Biological Conversion
Convert those raw nest counts into **Annual Nesters** and **Total Adult Females** using species-specific parameters.

```{r abundance, eval=FALSE}
# You can provide these values directly or answer the prompts
abundance <- calculate_abundance(
  clean_data, 
  clutch_freq = 5.5, 
  remig_int = 3.2
)
```

## Step 4: Trend Modeling & Visualization
Finally, run the Bayesian State-Space model. This will estimate the hidden population state while accounting for observation error.

```{r modeling, eval=FALSE}
# Run the model (use parallel = TRUE for faster results on large datasets)
results <- run_turtle_model(abundance, iter = 10000, parallel = FALSE)

# Generate the "Ghost Plot"
plot_turtle_status(results, abundance, species_name = "Leatherback")
```



## Conclusion
Once the plot is generated, check the **Annual Trend (U)** in the subtitle. A negative value indicates a population decline, while a positive value indicates growth.
