---
title: "Nestprepper Workflow Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nestprepper Workflow Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(DiagrammeR)
```

## How to use `nestprepper`

### Decision Flowchart

Select a path below to see the required steps.

```{r, echo=FALSE, fig.align='center'}
# 1. Create the diagram object
d <- mermaid("
graph TD
    Start((Start: I have<br/>Nesting Data)) --> Type{What is your<br/>data format?}
    
    Type -- Monthly Counts --> Pute[nestpute package<br/>Impute gaps & aggregate]
    Type -- Annual Totals --> Prep[nestprepper app<br/>QAQC & Bayesian Trend]
    
    Pute --> Prep
    
    Prep --> Output{What is your<br/>Goal?}
    
    Output -- Abundance --> Result1[Calculate Annual<br/>Nesting Females]
    Output -- Trend --> Result2[Bayesian State-Space<br/>Growth Rate %]
    
    style Start fill:#E1F5FE,stroke:#01579B,stroke-width:2px
    style Prep fill:#C8E6C9,stroke:#2E7D32,stroke-width:2px
    style Pute fill:#FFF9C4,stroke:#FBC02D,stroke-width:2px
    style Result1 fill:#F3E5F5,stroke:#7B1FA2
    style Result2 fill:#F3E5F5,stroke:#7B1FA2
")

# 2. Render it as a widget, but use widgetframe or standard output 
# to isolate it from the text.
d
```

---

## Data Paths

### 1. The `nestpute` Path (Under Development)
If you have raw monthly files (e.g., counts for January, February, March), you will eventually use the [nestpute](https://github.com/AnnaOrtega-NOAA/nestpute) package.
* **Process:** Impute monitoring gaps and aggregate into annual totals.
* **Current Status:** In development. Please aggregate data manually for now.

### 2. The `nestprepper` Path (Current)
If you already have one total count per year, you can proceed directly to the [nestprepper dashboard](https://github.com/AnnaOrtega-NOAA/nestprepper).
* **Process:** Interactive outlier detection, biological parameter conversion, and MCMC trend modeling.
* **Output:** Abundance estimates, Trend ($U$), and Ghost Plots for status reports.

---

### NOAA Open Science
By following this workflow, we ensure that every step from raw data to final trend is documented, reproducible, and standardized across different nesting regions.

> Ortega, A. (2026). nestprepper: An R package for sea turtle nesting data integration and trend modeling.
